---
title: "Part 1 - R introduction"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
  word_document: default
---

Simon says:

 - **learn is hard**: remember all the hours learning biology concepts;
 - **be patient and practise**: remember your experiments rarely success with your first attempt;
 - **use fresh eyes**: do not try to compare (yes, there is similarity with Excel or other); adopt the foreign language/culture learning process;
 - **be organised**: apply your experimental methodologies to your computational analysis; track your work with a *digital* **lab notebook**, e.g., with RMardown files. ;-)


In this introduction to R, we will present the main R concepts, by introducing
R variables, vectors, matrices, lists, functions and packages.

Please read http://adv-r.had.co.nz/Style.html.

# Variable

In R, we have two main types of values: numeric values and characters. Numerics:
```{r}
12
3.14
-42.12
```
and characters:
```{r}
"text"
"Rock and Roll"
```
The characters also named strings are between double-quote `"` (french keyboard: key 3).

We can save these values in variables,
```{r}

a <- 3.14

```
so we can easily access them later or re-use them:
```{r}

a + a
a

```

We can compute, save the value and display it:
```{r}

result <- 2*a + 3 * pi
result

```


### Exercises

1) Assign a number of your choice to the variable `b`
```{r}

```

2) Show what's in `b`
```{r}

```

3) Multiply `a` by `2`
```{r}

```

4) Add `10` to `a`
```{r}

```

5) Assign your name to an object called name
```{r}

```



# Vector / List

Vector contains several values:
```{r}

b <- 1:10

```

We can concatenate these values with the `c()` function:
```{r}

c(b, b)

```

We can also apply mathematical formulas to them:
```{r}

b + b
3 * b

```


### Exercises

6) Concatenate `a` and `b`
```{r}

```

7) Concatenate your name and `a`
```{r}

```

8) Add `a` to `b`
```{r}

```



# Built-in Function

Many functions exist in R, and greatly simplify our life.


How to compute the mean of the variable `b` ? The naive way:
```{r}

(1 + 2 + 3 + 4 + 5 + 6 + 7 + 8 + 9 + 10)/10

```
And now, what does it happen if `b <- 5:17`? We have to modify the formula.
Not enough powerful. R has built-in functions to easily compute common
quantities.

For example, the mean is computed by:
```{r}

mean(b)

```
Moreover, we can compose the function and the mathematical operations
(also fucntion).
```{r}

m <- mean(b + b)
m

```
What about the median? Easy:
```{r}

median(b)

```
And the number of elements (the length of the vector/list):
```{r}

length(b)
length(b + b)
length(c(b, b))

```
The concatenation `c()` is a built-in function.


Let sum all the elements, find the minimum, the maximum:
```{r}

sum(b)
min(b)
max(b)

```
Or print a statistical summary:
```{r}

summary(b)

```

List the 6 first elements:
```{r}

head(b)

```
or the 6 last:
```{r}

tail(b)

```
or the 4 last:
```{r}

tail(b, n=4)

```
Be careful, R does not warn even if we are asking *weird* stuff:
```{r}

head(b, n=20)

```

We can read help using:
```{r}

?tail

```
or the Help facilities of RStudio.



### Exercises


9) Create the vector 13, 13, 13, 13, 14, 14, 16, 18, 21 and asign it to a variable:
```{r}

```

10) Compute the median:
```{r}

```

11) Compute the mean:
```{r}

```

12) Is it normal? ;-)

13) Compute the length:
```{r}

```

14) Sum all the values:
```{r}

```

15) Compute the mean using the previous sum and length
```{r}

```


# Block (chunk) with comment

We sometimes want to insert a comment to remind the purpose of an
action.

A comment starts with the symbol sharp `#`.
```{r}

# this is a comment

```
and a comment does nothing when evaluated.

It is really useful to explain several lines:
```{r}

# Assign a list from 10 to 25 to the variable named var
var <- 10:25

# Select first elements
head(var)

var + var # return a new list

# Eval the length the concatenated list
length(c(var, var))

```

The comments do not appear in the Output block.
However they appear in the final document (see this doc).


# Matrices / Table

Matrices can be compared to tables: they are two dimensional and can contain numbers, characters, etc.

Let load a matrix:
```{r}
load("cyto.RData")
cyto
```

We can check the number of rows and columns in the cyto object:
```{r}

nrow(cyto)
ncol(cyto)
dim(cyto)

```

Load data from CSV file and show a summary:
```{r}

df <- read.csv2(file='example.csv', header=TRUE, dec=",", sep=";")
dim(df)
summary(df)

```


# Indexing

We can display the element in the first row and the first column, or the
2nd row and 3rd column.
```{r}

cyto[1, 1]
cyto[2, 3]

```
The indexing convention is: row, column.


All the elements of a specific column of a matrix can be selected by its
number, here the 3rd column:
```{r}

fcs_a <- cyto[, 3]
fcs_a

```

How to concatenate two columns ?
```{r}

# 1. select column 2 and column 5
# 2. concatenate them
# 3. store in the variable
fcs_b <- c(cyto[, 2], cyto[, 5])
fcs_b

# 1. concatenate the number 2 and 5
# 2. select the columns 2 and 5
# 3. store
fcs_c <- cyto[, c(2, 5)]
fcs_c

# Do you see the difference ?

```

The column can also be selected by typing its column name directly:
```{r}

cd45 <- cyto[, "CD34"]
cd45
cd45_b <- cyto[, 5]
cd45_b

```
Idem to select several columns:
```{r}

cd45_c <- cyto[, c("CD34", "CD45")]
cd45_c

```

However, you cannot mix number and name to select columns. 
This chunk returns an error:
```{r}

# cyto[, c(1, "CD34")]

```

Conversely, we can select a specific row, here the second:
```{r}
cell2 <- cyto[2, ]
cell2
```


## Important to remember

 - The names are important.
 - The spaces in column name are hell.

**Good names** of variables, columns, files, etc. **ease a lot** the
analysis, especially when things are not going as expected.


### Exercises

16) Show the 2 first rows of the cyto matrix
```{r}

```

17) Show the `FSC-A` value of the first cell
```{r}

```



# More on list

Lists in R can be compared to vectors, which can contain various objects:
```{r}

random_numbers <- c(1, 2, 3, 7)
random_letters <- c("A", "B", "C", "Z")

```
The variables `random_numbers` and `random_letters` are two lists
containing 4 numbers or 4 letters.

These two lists can be associated to create another list:
```{r}

my_list_b <- list(random_letters, random_numbers)
my_list_b

```
The variable `my_list_b` is a list containing two sub-lists. These
sub-lists can be named:
```{r}

my_list <- list(my_numbers = random_numbers,
                my_letters = random_letters)
my_list

```

The different items of a list can then be accessed by using the `$` and
typing their name directly:
```{r}

my_list$my_numbers

```


### Exercises

18) Create a list containing your name, and job:
```{r}

```


# Packages

Packages contain sets of useful functions, which can be installed, loaded, and
then used in R. A package needs to be installed only once on your computer:
```{r}

#install.packages("Rtsne")

```

But needs to be loaded every time you wish to use it:
```{r}

#library(Rtsne)

```

The functions of an installed package can then be used. They often need the user to set parameters:
```{r}

#tsne <- Rtsne(cyto)

```

This line would generate an error, we need to lower the perplexity as we only
have 10 cells.

```{r}

#tsne <- Rtsne(cyto, perplexity = 2)
#tsne$Y

```

You can always ask help to R, if you wish to know how a function works:
```{r}

#?Rtsne

```


# Define your own function

It is easy to define new functions. For example, I redefine my way to compute the mean:
```{r}

moyenne <- function(lst) {
  sumall <- sum(lst)
  len <- length(lst)
  moy <- sumall / len
  moy
}

```
Then I can use it:
```{r}

onelist <- 1:10
moyenne(onelist)

```
and the result is the same than `mean(onelist)` i.e., `r mean(onelist)`.

It is out of this presentation, but R is powerful and you can write loops over list, test value, etc.
```{r}

moyenne.absolute <- function(lst) {
  # init temporary variables
  nelem <- 0
  tot <- 0
  # loop over all the elements of the list
  for (val in lst) {
    if (val > 0 ) {
      tot <- tot + val
    } else {
      tot <- tot - val
    }
    nelem <- nelem + 1
  }
  # compute something
  moy <- tot / nelem
  # return it
  moy
}

```
and this function computes `mean(abs(lst))`. Let check that:
```{r}

one <- 1:10
onelist <- c(-one, 2*one)

moyenne.absolute(onelist) == mean(abs(onelist))

```
Well, if this section is not clear for you, then do not worry. We will talk again later.



# RMarkdown summary

 - list starts with `-`
 - *italics* is `*italics*`
 - **bold** is `**bold**`
 - `code` is `` `code` ``
 - equation $E=mc^2 + \sum x_i$ is `$E=mc^2 + \sum x_i$`
 - section starts with `#`, subsection with `##`, subsubsubsection with `###`, etc.
 - [hyperlink](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf) is
   `[hyperlink](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)`
 
## Subsection talking about table

You can also pretty print, e.g., a table:
```{r echo=FALSE, results='asis'}
library(knitr)
kable(cyto[3:7,c("CD45", "CD34", "CD56")], caption='Legend of the cyto table')
```

### Subsubsection talking about figure

```{r echo=FALSE, fig.cap="My nice caption of my nice figure"}
plot(cyto[,"FSC-A"], cyto[, "SSC-A"])
```


# Further readings

You could also have a look at the following online resources:

 - https://rmarkdown.rstudio.com/lesson-1.html
 - https://www.rstudio.com/online-learning/#r-programming
 - https://www.youtube.com/watch?v=o0Y478jOjGk (11min29)
 - https://www.youtube.com/watch?v=u1r5XTqrCTQ (2min52)
 
Please read http://adv-r.had.co.nz/Style.html.
